---
title: "Baseball Data Prep"
author: "James Richards"
date: "18/04/2022"
output: pdf_document
---

# Loading packages needed for data cleaning
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(janitor)
library(dplyr)
library(readxl)
library(kableExtra)
library(rstatix)
library(car)
library(caTools)
```

# Regular Season

## Loading in data
```{r, echo = FALSE}
#Installing data
Batting <- Batting
People <- People
Batting_post <- BattingPost
```


## Combining data
```{r}
Data <- left_join(People, Batting, by = "playerID")
```


## Calculating OPS and age
```{r}
Data$OBP <- ((Data$H + Data$BB + Data$HBP)/(Data$AB + Data$BB + Data$HBP + Data$SF))

Data$SLG <- ((Data$H + (2 * Data$`X2B`) + (3 * Data$`X3B`) + (4 * Data$HR))/(Data$AB))

Data$OPS <- Data$OBP + Data$SLG

Data$Age <- (Data$yearID - Data$birthYear)
```


## Removing NA and 0 for certain variables 
```{r}
Data <- Data[Data$AB >= 502, ]
Data <- Data[!is.na(Data$OPS),]
Data <- Data[!is.na(Data$Age),]
Data <- Data[!is.na(Data$weight),]
Data <- Data[!is.na(Data$height),]
Data <- Data[!is.na(Data$bats),]
Data <- Data[!is.na(Data$throws),]
```


## Saving players used
```{r}
player_id_list <- list(Data$playerID)
```


## Keeping data to be used
```{r}
Data = subset(Data, select = c("Age", "weight", "height", "bats", "throws", "OPS"))
```



# Post Season

## Combining data
```{r}
Data_post <- left_join(People, Batting_post, by = "playerID")
```


## Calculating OPS and age
```{r}
Data_post$OBP <- ((Data_post$H + Data_post$BB + Data_post$HBP)/(Data_post$AB + Data_post$BB + Data_post$HBP + Data_post$SF))

Data_post$SLG <- ((Data_post$H + (2 * Data_post$`X2B`) + (3 * Data_post$`X3B`) + (4 * Data_post$HR))/(Data_post$AB))

Data_post$OPS <- Data_post$OBP + Data_post$SLG

Data_post$Age <- (Data_post$yearID - Data_post$birthYear)
```


## Removing NA and 0 for certain variables 
```{r}
Data_post <- Data_post[! Data_post$playerID %in% player_id_list, ]
Data_post <- Data_post[!is.na(Data_post$OPS),]
Data_post <- Data_post[!is.na(Data_post$Age),]
Data_post <- Data_post[!is.na(Data_post$weight),]
Data_post <- Data_post[!is.na(Data_post$height),]
Data_post <- Data_post[!is.na(Data_post$bats),]
Data_post <- Data_post[!is.na(Data_post$throws),]
```


## Keeping data to be used
```{r}
Data_post = subset(Data_post, select = c("Age", "weight", "height", "bats", "throws", "OPS"))
```


# Model building

##Splitting data sets into test and training datasets
```{r, echo = FALSE}
set.seed(101)
#Regular season train and test dataset
dt_reg = sort(sample(nrow(Data), nrow(Data)*.5))
train_reg <- Data[dt_reg,]
test_reg <- Data[-dt_reg,]

train_reg <- data.frame(train_reg)
test_reg <- data.frame(test_reg)

#Post season train and test dataset
dt_post = sort(sample(nrow(Data_post), nrow(Data_post)*.5))
train_post <- Data_post[dt_post,]
test_post <- Data_post[-dt_post,]

train_post <- data.frame(train_post)
test_post <- data.frame(test_post)
```



##Creating base models 
```{r, echo = FALSE}
Base_Model_Reg <- lm(`OPS` ~ `Age` + `weight` + `height` + `bats` + `throws`, data = train_reg)

Base_Model_Post <- lm(`OPS` ~ `Age` + `weight` + `height` + `bats` + `throws`, data = train_post)
```



##Checking that Checking conditions I & II and assumptions are met
```{r, echo = FALSE}
pairs(train_reg[, c("weight", "height", "OPS")])
pairs(train_post[, c("weight", "height", "OPS")])
```


```{r, echo = FALSE}
plot(train_reg$`OPS` ~ fitted(Base_Model_Reg), main = "Y versus Y-hat regular season", xlab = "Y-hat", ylab = "Y")
abline(a = 0, b = 1)
lines(lowess(train_reg$`OPS` ~ fitted(Base_Model_Reg)), lty = 2)

plot(train_post$`OPS` ~ fitted(Base_Model_Post), main = "Y versus Y-hat post season", xlab = "Y-hat", ylab = "Y")
abline(a = 0, b = 1)
lines(lowess(train_post$`OPS` ~ fitted(Base_Model_Post)), lty = 2)
```


```{r, echo = FALSE}
##Q-Q Plot for regular season training dataset
r_reg <- resid(Base_Model_Reg)
plot(r_reg ~ fitted(Base_Model_Reg), main = "Res vs. Fitted", xlab = "Fitted", ylab = "Residuals")
plot(r_reg ~ train_reg$`weight`)
plot(r_reg ~ train_reg$`height`)
plot(r_reg ~ train_reg$`OPS`)

qqnorm(r_reg)
qqline(r_reg)
```


```{r, echo = FALSE}
##Q-Q Plot for post season training dataset
r_post <- resid(Base_Model_Post)
plot(r_post ~ fitted(Base_Model_Post), main = "Res vs. Fitted", xlab = "Fitted", ylab = "Residuals")
plot(r_post ~ train_post$`weight`)
plot(r_post ~ train_post$`height`)
plot(r_post ~ train_post$`OPS`)

qqnorm(r_post)
qqline(r_post)
```



##Transforming variables to satisfy assumptions if needed
```{r, echo = TRUE}
##Box-Cox transformation for regular season
boxCox(Base_Model_Reg, family = "bcnPower")

##Box-Cox transformation for post season
boxCox(Base_Model_Post, family = "bcnPower")
```


```{r, echo = TRUE}
##Box-Cox transformation for quantitative predictor variable regular season
p_reg <- powerTransform(cbind(train_reg[,1], train_reg[,2], train_reg[,3]) ~ 1, family = "bcnPower")
summary(p_reg)

p_post <- powerTransform(cbind(train_post[,1], train_post[,2], train_post[,3]) ~ 1, family = "bcnPower")
summary(p_post)
```


```{r, echo = FALSE}
##Regular Season
##Transforming Y variable 
train_reg$OPS <- (train_reg$OPS)

##Transforming X variables
train_reg$Age <- (train_reg$Age)^(-0.5)
train_reg$weight <- (train_reg$weight)^(-1)
train_reg$height <- (train_reg$height)


#Post Season
##Transforming Y variable 
train_post$OPS <- (train_post$OPS)^(0.5)

##Transforming X variables
train_post$Age <- (train_post$Age)
train_post$weight <- (train_post$weight)^(-1)
train_post$height <- (train_post$height)^(2)
```


##Checking for multicollinearity
```{r, echo = FALSE}
##Using VIF function to check for multicollinearity
vif(Base_Model_Reg)

vif(Base_Model_Post)
```


```{r, echo = FALSE, fig.width=48, fig.height=27}
##Determining which predictors have perfect one-to-one relationships
pairs(train_reg[, c("Age", "weight", "height")])

pairs(train_post[, c("Age", "weight", "height")])
```


```{r, echo = FALSE}
##Creating correlation matrix for data frames in regular season
train_numeric_reg = subset(train_reg, select = c(Age, weight, height))
train_numeric <- data.frame(train_numeric_reg)
cor(train_numeric_reg)

##Creating correlation matrix for data frames in post season
train_numeric_post = subset(train_post, select = c(Age, weight, height))
train_numeric <- data.frame(train_numeric_post)
cor(train_numeric_post)
```


##Performing t-test
```{r, echo = FALSE}
##t-test on finalized base model for regular season
summary(Base_Model_Reg)

##t-test on finalized base model for post season
summary(Base_Model_Post)
```


##creating reduced models
```{r, echo = FALSE}
New_Model_Reg <- lm(`OPS` ~ `Age` + `weight` + `height` + `bats`, data = train_reg)

New_Model_Post <- lm(`OPS` ~ `weight` + `height`, data = train_post)
```


##partial f-tests between base and reduced models
```{r, echo = FALSE}
#Test between regular season models
anova(New_Model_Reg, Base_Model_Reg)

#Test between post season models
anova(New_Model_Post, Base_Model_Post)
```


##Models
```{r, echo = FALSE}
New_Model_Reg <- lm(`OPS` ~ `Age` + `weight` + `height` + `bats`, data = train_reg)

New_Model_Post <- lm(`OPS` ~ `weight` + `height`, data = train_post)
```


##Performing t-test again
```{r, echo = FALSE}
##t-test on finalized base model for regular season
summary(New_Model_Reg)

##t-test on finalized base model for post season
summary(New_Model_Post)
```



# Creating test models
```{r, echo = FALSE}
#Transforming variables indicated by box cox to satisfy conditions I & II

#Regular Season
##Transforming Y variable 
test_reg$OPS <- (test_reg$OPS)

##Transforming X variables
test_reg$Age <- (test_reg$Age)^(-0.5)
test_reg$weight <- (test_reg$weight)^(-1)
test_reg$height <- (test_reg$height)


#Post Season
##Transforming Y variable 
test_post$OPS <- (test_post$OPS)^(0.5)

##Transforming X variables
test_post$Age <- (test_post$Age)
test_post$weight <- (test_post$weight)^(-1)
test_post$height <- (test_post$height)^(2)
```

```{r, echo = FALSE}
Test_Model_Reg <- lm(`OPS` ~ `Age` + `weight` + `height` + `bats`, data = test_reg)

Test_Model_Post <- lm(`OPS` ~ `weight` + `height`, data = test_post)
```


# Looking at fit of models

##Final Model fit in training dataset
```{r, echo = FALSE}
#Regular season final model
plot(train_reg$`OPS` ~ fitted(New_Model_Reg), main = "Y versus Y-hat", xlab = "Y-hat", ylab = "Y")
abline(a = 0, b = 1)
lines(lowess(train_reg$`OPS` ~ fitted(New_Model_Reg)), lty = 2)

#Post season final model
plot(train_post$`OPS` ~ fitted(New_Model_Post), main = "Y versus Y-hat", xlab = "Y-hat", ylab = "Y")
abline(a = 0, b = 1)
lines(lowess(train_post$`OPS` ~ fitted(New_Model_Post)), lty = 2)
```


```{r, echo = FALSE}
#Regular season
rr_reg <- resid(New_Model_Reg)
plot(rr_reg ~ fitted(New_Model_Reg), main = "Res vs. Fitted", xlab = "Fitted", ylab = "Residuals")

qqnorm(rr_reg)
qqline(rr_reg)

#Post season
rr_post <- resid(New_Model_Post)
plot(rr_post ~ fitted(New_Model_Post), main = "Res vs. Fitted", xlab = "Fitted", ylab = "Residuals")

qqnorm(rr_post)
qqline(rr_post)
```


##Final Model fit in testing dataset
```{r, echo = FALSE}
#Regular season final model
plot(test_reg$`OPS` ~ fitted(Test_Model_Reg), main = "Y versus Y-hat", xlab = "Y-hat", ylab = "Y")
abline(a = 0, b = 1)
lines(lowess(test_reg$`OPS` ~ fitted(Test_Model_Reg)), lty = 2)

#Post season final model
plot(test_post$`OPS` ~ fitted(Test_Model_Post), main = "Y versus Y-hat", xlab = "Y-hat", ylab = "Y")
abline(a = 0, b = 1)
lines(lowess(test_post$`OPS` ~ fitted(Test_Model_Post)), lty = 2)
```


```{r, echo = FALSE}
#Regular season
t_reg <- resid(Test_Model_Reg)
plot(t_reg ~ fitted(Test_Model_Reg), main = "Res vs. Fitted", xlab = "Fitted", ylab = "Residuals")

qqnorm(t_reg)
qqline(t_reg)

#Post season
t_post <- resid(Test_Model_Post)
plot(t_post ~ fitted(Test_Model_Post), main = "Res vs. Fitted", xlab = "Fitted", ylab = "Residuals")

qqnorm(t_post)
qqline(t_post)
```


# Summaries of final models
```{r, echo = FALSE}
##t-test on finalized base model for regular season
summary(Test_Model_Reg)

##t-test on finalized base model for post season
summary(Test_Model_Post)
```